Robustness of MinD oscillation in \emph{Escherichia coli} with diverse cell shapes

### 2013-11-21 paper/paper.tex
modified the model section and the shapes section in the paper Tested everything on node37.

\newcommand{\fixme}[1]{\red{[#1]}}
\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\renesays}[1]{{\color{red} [\blue{Rene:} \emph{#1}]}}

@@

\section{Model and Methods}
We implement the model of Huang \emph{et al.}  for the behavior of the
MinD and MinE proteins inside the cell~\cite{huang2003dynamic}.
Figure~\ref{fig:reactions} shows the reaction process.  The
cytoplasmic MinD:ADP complex undergoes nucleotide exchange and is
changed into the MinD:ATP complex.  This will naturally diffuse
towards and attach to the cell membrane.  A cytoplasmic MinE will
attach to the wall bound MinD:ATP complex and after a time will
activate ATP hydrolosis.  This breaks up the complex, releasing MinE,
phosphate, and MinD:ADP back into the cell.  The MinD:ADP will undergo
nucleotide exchange and begin again the cyclic process.  There is a
mechanism for MinD:ATP clustering that is shown in
Equation~\ref{eq:d-on-wall}.  When brushing up against the cell
membrane, the MinD:ATP attachement rate increases linearly with the
density of MinD:ATP already attached.  This model is defined by a set of
five reaction-diffusion equations:

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE},\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP},
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E,
\end{multline}
\begin{multline}
  \frac{\partial \sigma_d}{\partial t} = -k_E\rho_d\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP},
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE},
\end{multline}

@@

Our diffusion and reaction rates are shown below.  We are interested
primarily with the effect of cellular shape on the protein
oscillations, so we follow Huang's work\cite{huang2003dynamic} and do
not deviate from these values.  Huang's simulations use total MinD and
MinE concentrations of $1,000/\mu m$ and $350/\mu m$, respectively, in
a cylindrical cell of radius $0.5\mu m$, and in our (non-cylindrical)
cells we use the same number of proteins per unit volume.

@@

We start our cells with the MinD and MinE concentrations that are
reported as wild type concentrations by Huang et al. These
concentration values are 1273 $\mu m^{-3}$ and 445 $\mu m^{-3}$,
respectively. A 3d grid is then constructed in cartesian coordinates,
with a grid spacing of .05\micron. We define our cell shapes and solve
the reaction-diffusion equations numerically to observe the time
evolution of the MinD and MinE concentrations inside the
cell. \renesays{This last sentence sweeps a lot under the rug. How
  much more detail should be given to the computational method?}

@@

In order to address the limits at which the stability of the
oscillations of the Min system break down, as observed by Mannik
\emph{et al.}~\cite{mannik2012robustness}, we model the Min system in
a variety of cell shapes.  We begin with natural pill-shaped cells,
and the proceed to examine a number of flattened pancake-like shapes,
which reflect the experiment of Mannik \emph{et al.}, in which
bacteria were confined within a thin slit~\cite{mannik2012robustness}.
Our pill shapes differ from those of Huang \emph{et al.} in that they
are cylinders with hemisphere endcaps instead of pure cylindridrical
shapes.  Our
cylinderidrical radius is $0.50\micron$ and the lengths of our cells
(measured between the tips of the endcaps) are $5\micron$, $4\micron$, $3\micron$,
and $2.5\micron$.

@@

Our first group of (three) flattened cell shapes are triangular.  One
of them is equilateral, one has a 3-4-5 right angle shape, and the
last is iscoscolese.

Our second group of (four) flattened cells are our own creations,
designed to investigate different patterns of protein oscillation
behavoir.

The cell shapes and the sections that we split them into in our data
analysis are shown in Figure \fixme{We need a big figure with all of this!!}.

Kubitschek has shown in multiple experiments that at the time of cell
division cells have a volume that is within a range of roughly
$1\micron^3$ to $2\micron^3$~\cite{kubitschek1990cell,
  kubitschek1968linear}.  We follow Huang's
simulations\cite{huang2003dynamic} and Mannik's experiments
\fixme{this link doesn't work = cite{mannik2012robustness}} and model
cells that are slightly larger than this range.  Amoung our flattened
cells the two dimensional length scales are tuned so that every total
cell volume is very close to $3\micron^3$.  \fixme{should we just
  leave it to the reader to figure out for themselves what our pill
  shape volumes are?}

To interpret the results, we generated several different plot views of
the printed simulation data. These plots included a time averaged view
of the protein densities in the cell; a plot tracking the location of
protein concentrations that were global maxima in space and local maxima in
time; and an animated view that showed the actual dispersion of
protein concentrations in the cell over time.
\section{Specific Results}

@@

The reaction shown in Figure~\ref{fig:reactions} is cyclical so we can
begin our observations at any stage in the cycle (which is presented
for the natural cell shape in Fig.~\ref{image-p} below).  We consider
first MinD:ATP floating freely in the cytoplasm.  The protein will not
react with any other floating protein, and it will not spontenously
shift back into the MinD:ADP stage of the cycle.  It will diffuse
until it attaches to a cell membrane.  This is will happen most
rapidly in regions of high curvature, where diffusion is more likely
to lead to an encounter with a wall.  This effect is made more
pronounced by the MinD:ATP tendency to attract and cluster on the
membrane, as reflected in Equation~\ref{eq:d-on-wall}, which shows
that when adjacent to the membrane, the rate of binding to the
membrane is porportional to the amount of MinD already bound to the
wall.  Initially, then, there is a strong tendency of MinD to cluster
on the polar regions of the cell, where the curvature is highest.

We next see the formation of MinE `rings' on the membranes of the
polar caps.  As the freely diffusing MinE enters the cap portion it
begins to encounter concentrated membrane bound MinD.  The MinE binds
to this MinD and remians for a time on the walls in compound form.
When it is released, it can diffuse naturally outward or inward
towards the cell.  If it diffuses inward, there is less cross
sectional wall for it to bump into and also less concentrated MinD:ATP
to react with.  If it diffuses outward, the curvature of the cell wall
cuases there to be a greater chance that it will encounter the cell
membrane (for the same reason outlined for the MinD:ATP above) and if
there is MinD:ATP bound there, there is a chance that it will bind to
that protein and react again.  If it does diffuse in this directin, it
will likely react with MinD:ATP before getting very far into the
encap.  Thus after dettaching from the membrane at the edge of the
MinD:ATP encap cluster, the MinE is most likely to travel a small
distance further into the polar region and then attach and react again
with MinD:ATP protein there.  This process results in the observed
qualitative 'encroaching MinE ring' behavoir that releases MinD:ADP
back into the cytoplasm.  This MinD:ADP will diffuse for a time before
spontanosly uindergowing hydrolysis \fixme{is this right?} and
reaching the MinD:ATP state that can bind to the membrane again.
There is a strong chance that it has by this time diffused out of the
polar region and is on its way to the opposite polar region of the
cell.  If it hasn't then it will reattach to the MinD:ATP on the cell
wall deeper into the endcap and will eventually be converged upon by
the MinE ring, only to be released again in MinD:ADP form.

Overall this process results in an extremely robust oscillatory
behavoir of polar selection and oscillation in a wide variety of cell
shapes.

>>>>

\newcommand{\fixme}[1]{\red{[#1]}}
\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\renesays}[1]{{\color{red} [\blue{Rene:} \emph{#1}]}}
\newcommand{\jeffsays}[1]{{\color{red} [\blue{Jeff:} \emph{#1}]}}

@@

\section{Model and Methods}
We implement the model of Huang \emph{et al.}  for the behavior of the
MinD and MinE proteins inside the cell~\cite{huang2003dynamic}.
Figure~\ref{fig:reactions} shows the reaction process.  The
cytoplasmic MinD:ADP complex undergoes nucleotide exchange and is
changed into the MinD:ATP complex.  This will naturally diffuse
towards and attach to the cell membrane.  A cytoplasmic MinE will
attach to the wall bound MinD:ATP complex and after a time will
activate ATP hydrolosis.  This breaks up the complex, releasing MinE,
phosphate, and MinD:ADP back into the cell.  The MinD:ADP will undergo
nucleotide exchange and begin again the cyclic process.  This model is
defined by a set of five reaction-diffusion equations:

\begin{multline}
  \frac{\partial \rho_{D:ADP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ADP}-k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  +\delta(d_w)k_{de}\sigma_{DE},\hspace{3.4cm}
\end{multline}
\begin{multline}
  \frac{\partial \rho_{D:ATP}}{\partial t} = \mathcal{D}_D\nabla^2\rho_{D:ATP}+k_D^{ADP\rightarrow ATP}\rho_{D:ADP}\\
  -\delta(d_w)[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
\end{multline}
\begin{multline}
  \frac{\partial \rho_E}{\partial t} = \mathcal{D}_E\nabla^2\rho_E+\delta(d_w)k_{de}\sigma_{DE}
  -\delta(d_w)k_E \sigma_D \rho_E
\end{multline}
\begin{multline}
  \frac{\partial \sigma_d}{\partial t} = -k_E\rho_d\rho_E
  +[k_D+k_{dD}(\sigma_D+\sigma_{DE})]\rho_{D:ATP}
  \label{eq:d-on-wall}
\end{multline}
\begin{multline}
  \frac{\partial \sigma_{DE}}{\partial t} = -k_{de}\sigma_{DE}+k_E\sigma_D\rho_E\hspace{3cm}
  \label{eq:FifthPDE}
\end{multline}

@@

Our diffusion and reaction rates are shown below.  We are interested
primarily in the effect of cellular shape on the protein
oscillations, so we follow Huang's work\cite{huang2003dynamic} and do
not deviate from the wild type values used in that work.  Huang's
simulations use total MinD and MinE concentrations of $1,000/\mu m$
and $350/\mu m$, respectively, in a cylindrical cell of radius $0.5\mu
m$, and in our (non-cylindrical) cells we use the same number of
proteins per unit volume.  These concentration values are 1273 $\mu
m^{-3}$ and 445 $\mu m^{-3}$, respectively.

@@

A 3d grid is then constructed in cartesian coordinates, with a grid
spacing of .05\micron. We define our cell shapes and solve the
reaction-diffusion equations numerically to observe the time evolution
of the MinD and MinE concentrations inside the cell. \renesays{This
  last sentence sweeps a lot under the rug. How much more detail
  should be given to the computational method?}

The reactions shown in Figure~\ref{fig:reactions} are cyclical so we
can begin our observations at any stage in the cycle (which is
presented for the natural, pill shapes cell in Fig.~\ref{image-p}
below).  We consider first MinD:ATP floating freely in the cytoplasm.
The protein will not react with any other floating protein, and it
will not spontenously shift back into the MinD:ADP stage of the cycle.
It will diffuse until it attaches to a cell membrane.  This will
happen most rapidly in cytoplasmic regions that are close to membrane
regions of high curvature, where diffusion in a random direction is
more likely to lead to an encounter with a wall.  This effect is made
more pronounced by the MinD:ATP tendency to attract and cluster on the
membrane, as reflected in Equation~\ref{eq:d-on-wall}, which shows
that when adjacent to the membrane, the rate of binding to the
membrane is porportional to the amount of MinD already bound to the
wall.  Initially, then, there is a strong tendency of MinD to cluster
on the polar regions of the cell, where the curvature is highest.

We next see the formation of MinE `rings' on the membranes of the
polar caps.  As the freely diffusing MinE enters the cap portion it
begins to encounter concentrated membrane bound MinD.  The MinE binds
to this MinD and remains for a time on the walls in compound form.
When it is released, it has the two options of diffusing naturally
outward or inward towards the cell.  If it diffuses inward, there is
less cross sectional wall for it to bump into and also less
concentrated MinD:ATP to react with.  If it diffuses outward, the
curvature of the cell wall cuases there to be a greater probability
per unit time that it will encounter the cell membrane (for the same
reason outlined for the MinD:ATP above) where there is more membrane
bound MinD:ATP. If it does diffuse in this directin, it will likely
react with MinD:ATP before getting very far into the encap.  This
process results in the observed qualitative 'encroaching MinE ring'
behavoir that releases MinD:ADP back into the cytoplasm.  This
MinD:ADP will diffuse for a time before spontanosly undergowing
nucleotide exchange and reaching the MinD:ATP state that
can bind to the membrane again.  There is a strong chance that it has
by this time diffused out of the polar region and is on its way to the
opposite polar region of the cell.  If it hasn't then it will reattach
to the MinD:ATP on the cell wall deeper into the endcap and will
eventually be converged upon by the MinE ring, only to be released
again in MinD:ADP form.

Overall this process results in extremely robust oscillatory
behavoir of polar selection and oscillation in a wide variety of cell
shapes.

@@

We mean to investigate the geometric limits of the Min system
oscillations, as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we model the Min system in a
variety of cell shapes.  We begin with naturally occuring pill-shaped
cells and then proceed to examine a number of flattened pancake-like
shapes, which reflect the experiment of Mannik \emph{et al.}, in which
bacteria were confined within a thin slit~\cite{mannik2012robustness}.
Our pill shapes differ from those of Huang \emph{et al.} in that they
are cylinders with hemisphere endcaps instead of pure cylindrical
shapes.  Our cylinderidrical radius is $0.5\micron$ and the lengths
of our cells (measured between the tips of the endcaps) are
$5\micron$, $4\micron$, $3\micron$, and $2.5\micron$.

@@

Our first group of (three) flattened cell shapes are triangular when
viewed from above.  One of them is equilateral, one has a 3-4-5 right
angle shape, and the last is iscoscolese.

Our second group of (two) flattened cells are designed to replicate
those published by Mannik, in order to compare the results of the
simulation model with the experimental results in that work.

Our third group of (three) flattened cells are our own creations,
designed to investigate different patterns of protein oscillation
behavoir.

Kubitschek has shown in multiple experiments that at the time of cell
division cells have a volume that is within a range of roughly
$1\micron^3$ to $2\micron^3$~\cite{kubitschek1990cell,
  kubitschek1968linear}.  We follow Huang's
simulations\cite{huang2003dynamic} and Mannik's experiments and model
cells that are slightly larger than this range.  Amoung our flattened
cells the two dimensional length scales are tuned so that total cell
volumes are very close to $3\micron^3$. \jeffsays{I'm not sure how
  true this is.  We should go through and check.  There've been a few
  implementations of printouts reporting volume in the cpp}
\section{Specific Results}

@@
